---
title: "cleaning_grouping"
format: html
code-fold: true
code-tools: true
editor_options: 
  chunk_output_type: console
---

# Grouping Items into Variables

```{r}

setwd("~/Desktop/R_projects/msc")


rm(list = ls()) # Clear Global Environment


library(tidyverse)
library(psych)
library(lavaan)
```

```{r}
data <- read_csv("data/middle_school_music_mindsets_clean_3_imputed.csv")
```

### Reclassify data as factor

```{r}
reclassFactor <- c("grade","school","race","hispanic_latino","gender","currentSchoolMusic","familyMusic")

data[reclassFactor] <- lapply(data[reclassFactor], as.factor)
```

### School Music Enrollment

```{r}
#Create binary variable for enrollment in music electives
data <- data %>% 
  mutate(school_music_elective = ifelse(currentSchoolMusic == "No music", "No",
                                        ifelse(is.na(currentSchoolMusic), NA, "Yes")))
```

### Subsets

Store demographic data, measurement data, and other character data (i.e, instrument, familyMusic) into separate variables.

```{r}

demographic_data <- data %>% 
  select(survey_id:currentSchoolMusic, school_music_elective)

measures_data <- data %>%
  select(s1_1:s3_28) %>%
  select_if(is.numeric)

instrument <- data %>% 
  select(instrument)

familyMusic <- data %>% 
  select(familyMusic)
```

### Reverse required items

1 = natural

-1 = reversed

```{r}
key <- c(1, 1, 1, 1, -1, 1, 1, 1, 1, 1, -1, 1, -1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, -1, 1, -1, 1, 1, 1, 1, -1, 1, 1, 1, 1, 1, 1, 1)

measures_data_reversed <- psych::reverse.code(keys = key, measures_data)

measures_data_reversed <- as.data.frame(measures_data_reversed)

measures_data_reversed <- measures_data_reversed %>% 
  rename(s1_5r = `s1_5-`,
         s1_11r = `s1_11-`,
         s1_13r = `s1_13-`,
         s3_14r = `s3_14-`,
         s3_16r = `s3_16-`,
         s3_21r = `s3_21-`)

write_csv(measures_data_reversed, "middle_school_music_mindsets_clean_measures_data_reversed_raw_scores.csv")

```

### Z Scores

```{r}
scaled_measures_data <- scale(measures_data_reversed) #z-scores

scaled_measures_data <- as.data.frame(scaled_measures_data) %>% #convert matrix to data frame
  rename_all(~ paste0(., "_z")) #rename column names to include "_z" to differentiate from raw scores later in the full data set
```

### Combine all subsets

```{r}
data <- cbind(demographic_data, instrument, familyMusic,measures_data_reversed, scaled_measures_data) #all original variables joined together and in new order.

```

### Adding Items to Latent Variables

I created two versions of the latent variables, one with raw scores, the other with z-scores (note: active_engagement and musical_training contain items from different scales i.e., 1-4 and 1-7, I only used a z-score version on those variables when comparing to other variables).

#### Gold-MSI: Active Engagement

```{r}
active_engagement <- data %>%
  mutate(active_engagement = s1_1 + s1_2 + s1_4 + s1_7 + s1_9 + s1_11r + s1_12 + s2_1 + s2_2) %>%
  select(active_engagement)
```

#### Gold-MSI: Active Engagement (z-score)

```{r}
active_engagement_z <- data %>%
  mutate(active_engagement_z = s1_1_z + s1_2_z + s1_4_z + s1_7_z + s1_9_z + s1_11r_z + s1_12_z + s2_1_z + s2_2_z) %>%
  select(active_engagement_z)
```

#### Gold-MSI: Musical Training

```{r}
musical_training <- data %>%
  mutate(musical_training = s1_5r + s1_13r + s2_3 + s2_4 + s2_5 + s2_6) %>%
  select(musical_training)
```

#### Gold-MSI: Musical Training (z-score)

```{r}
musical_training_z <- data %>%
  mutate(musical_training_z = s1_5r_z + s1_13r_z + s2_3_z + s2_4_z + s2_5_z + s2_6_z) %>%
  select(musical_training_z)
```

#### Growth Mindset

```{r}
growth_mindset <- data %>%
  mutate(growth_mindset = s1_3 + s1_6 + s1_8 + s1_10) %>%
  select(growth_mindset)
```

#### Growth Mindset (z-score)

```{r}
growth_mindset_z <- data %>%
  mutate(growth_mindset_z = s1_3_z + s1_6_z + s1_8_z + s1_10_z) %>%
  select(growth_mindset_z)
```

#### Mood Management (Music Self Concept)

```{r}
mood_management_msc <- data %>%
  mutate(mood_management_msc = s3_1 + s3_3 + s3_13 + s3_17 + s3_19 + s3_22) %>%
  select(mood_management_msc)

```

#### Mood Management (Music Self Concept; z-score)

```{r}
mood_management_msc_z <- data %>%
  mutate(mood_management_msc_z = s3_1_z + s3_3_z + s3_13_z + s3_17_z + s3_19_z + s3_22_z) %>%
  select(mood_management_msc_z)

```

#### Community (Music Self Concept)

```{r}
community_msc <- data %>%
  mutate(community_msc = s3_2 + s3_15 + s3_23 + s3_25) %>%
  select(community_msc)
```

#### Community (Music Self Concept; z-score)

```{r}
community_msc_z <- data %>%
  mutate(community_msc_z = s3_2_z + s3_15_z + s3_23_z + s3_25_z) %>%
  select(community_msc_z)
```

#### Musical Ability (Music Self Concept)

```{r}
musical_ability_msc <- data %>%
  mutate(musical_ability_msc = s3_4 + s3_6 + s3_8 + s3_14r + s3_16r) %>%
  select(musical_ability_msc)
```

#### Musical Ability (Music Self Concept; z-score)

```{r}
musical_ability_msc_z <- data %>%
  mutate(musical_ability_msc_z = s3_4_z + s3_6_z + s3_8_z + s3_14r_z + s3_16r_z) %>%
  select(musical_ability_msc_z)
```

#### Adaptive Self (Music Self Concept)

```{r}
adaptive_musical_self_msc <- data %>%
  mutate(adaptive_musical_self_msc = s3_5 + s3_10 + s3_24 + s3_28) %>%
  select(adaptive_musical_self_msc)
```

#### Adaptive Self (Music Self Concept; z-score)

```{r}
adaptive_musical_self_msc_z <- data %>%
  mutate(adaptive_musical_self_msc_z = s3_5_z + s3_10_z + s3_24_z + s3_28_z) %>%
  select(adaptive_musical_self_msc_z)
```

#### Ideal Self (Music Self Concept)

```{r}
ideal_musical_self_msc <- data %>%
  mutate(ideal_musical_self_msc = s3_7 + s3_11 + s3_18 + s3_20 + s3_27) %>%
  select(ideal_musical_self_msc)
```

#### Ideal Self (Music Self Concept; z-score)

```{r}
ideal_musical_self_msc_z <- data %>%
  mutate(ideal_musical_self_msc_z = s3_7_z + s3_11_z + s3_18_z + s3_20_z + s3_27_z) %>%
  select(ideal_musical_self_msc_z)
```

#### Movement & Dance (Music Self Concept)

```{r}
movement_dance_msc <- data %>%
  mutate(movement_dance_msc = s3_9 + s3_12 + s3_21r + s3_26) %>%
  select(movement_dance_msc)
```

#### Movement & Dance (Music Self Concept; z-score)

```{r}
movement_dance_msc_z <- data %>%
  mutate(movement_dance_msc_z = s3_9_z + s3_12_z + s3_21r_z + s3_26_z) %>%
  select(movement_dance_msc_z)
```

#### Music Self-Concept (Full Scale)

```{r}

msc_full <- mood_management_msc	+ 
  community_msc	+ 
  musical_ability_msc	+ 
  adaptive_musical_self_msc + 
  ideal_musical_self_msc + 
  movement_dance_msc

colnames(msc_full) <- "msc_full"

```

#### Music Self-Concept (Full Scale; z-scores)

```{r}

msc_full_z <- scale(msc_full)

colnames(msc_full_z) <- "msc_full_z"

describe(msc_full_z)

```

## Add all Variables to full dataset

```{r}

data <- cbind(data,
               growth_mindset,
               active_engagement,
               musical_training,
               mood_management_msc,
               community_msc,
               musical_ability_msc,
               adaptive_musical_self_msc,
               ideal_musical_self_msc,
               movement_dance_msc,
               msc_full, 
               active_engagement_z,
               musical_training_z,
               growth_mindset_z,
               mood_management_msc_z,
               community_msc_z,
               musical_ability_msc_z,
               adaptive_musical_self_msc_z,
               ideal_musical_self_msc_z,
               movement_dance_msc_z,
               msc_full_z)
```

## Music No Ability

To examine the difference between the musical ability sub-scale and the composite msc score, I created a variable that represents a composite of all sub-scales but excludes the musical ability sub-scale, thus representing a third outcome.

```{r}

#create MSC Full without musical ability subscale
msc_no_ability <- data$mood_management_msc +
  data$community_msc +
  data$adaptive_musical_self_msc +
  data$ideal_musical_self_msc +
  data$mood_management_msc

data <- cbind(data, msc_no_ability)
```

## Music No Ability (z-scores)

```{r}
msc_no_ability_z <- scale(msc_no_ability)
data <- cbind(data, msc_no_ability_z)
```

## Subset for Data Analysis (does not contain individual items for measures related to latent variables)

```{r}

data_partial <- data %>% 
  select(!matches("s1|s2|s3"))
```

```{r}

write_csv(data, "data/middle_school_music_mindsets_clean_full.csv")

write_csv(data_partial, "data/middle_school_music_mindsets_clean_partial.csv")
```
